---
title: "PCA intro"
author: "kp"
date: "2022-10-31"
output: 
    ioslides_presentation: 
      css: 'scrollable_slides.css'

---

```{r setup, include=FALSE}
```

## PCA first example {.build}

made up values of height and weight 

```{r, echo=T}

height = rnorm(100, m=5, s=1)
weight = height*34+rnorm(100, 0,5)
df = data.frame(height, weight)
df$names = NA
df[which.max(df$height), "names"] = "tallest"
df[which.min(df$height), "names"] = "shortest"
plot(df[,1:2])
text(df[,1:2], df$names)

```


## Normalize data to make it easier to work with  {.build}

```{r, echo=T}
#center and scale data
df$height = df$height-mean(df$height)
df$weight = df$weight-mean(df$weight)
df$height=df$height/sd(df$height)
df$weight=df$weight/sd(df$weight)
plot(df[,1:2]) 
text(df[,1:2], df$names)


```

## compute principal components {.build}

```{r, echo=T}
#compute principle components
pca = prcomp(df[,1:2], scale=F)

str(pca)

```

- rotation: shows the loadings of the variables in each PC.  
- x:  the new values of the samples across the PCs.  
- sdev:  shows how much variance is explained by each PC.




## Examine first principal component (eigenvector)

```{r, echo=T}

rot = pca$rotation

#which direction maximized variance?
plot(df[,1:2]) 
text(df[,1:2], df$names)
lines(rbind(c(0,0),rot[,1]), col="red", lwd=3)

```

## examine scores from PC1 (eigenvalues)

```{r, echo=T}
#plot samples along the first principal component
plot(pca$x[,1], rep(0, nrow(df)))
text(pca$x[,1], rep(0, nrow(df)), df$names)

```


## PCA second example {.build}


```{r, echo=T}
mtcars

```

## PCA compute and plot {.build}

```{r, echo=T}
pca2 = prcomp(mtcars)

plot(pca2)

```

## PCA compute and plot {.build}

```{r, echo=T}
plot(pca2$x[,1:2])
text(pca2$x[,1:2], rownames(mtcars))

```

## Check out the PC loadings {.build}

these show how much each variable is contributing to the PC.


```{r, echo=T}
pca2$rotation

```


## another way to plot PCA {.build}



```{r, echo=T}
library(ggfortify)
autoplot(pca2, colour="disp", label=T)
autoplot(pca2, colour="hp", label=T)
autoplot(pca2, label=T, loadings=T, loadings.label=T)

```
